---
author: "Yu Chen, Zifei Dong, Ning Pan, Sifan Tao, Yao Yao"
title: "565 Final Project House Price EDA"
date: "2023-03-07"
output: html_document
---


House Prices Prediction
===

We take will use the [House Prices](https://www.kaggle.com/competitions/house-prices-advanced-regression-techniques/data) data from Kaggle. The measured variables can be seen [here](https://agaresd47.github.io/stor565_final_project.github.io/House%20Price%20Prediction/Data%20Description.pdf)
```{r, eval=TRUE, include= FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load the required library
library(ggplot2)
library(dplyr)
library(corrplot)
```




Exploratory data analysis：
===

```{r}
# Read the train.csv file into a data frame
house_prices <- read.csv("train.csv", header = TRUE, stringsAsFactors = FALSE)

dim(house_prices)
```


```{r}
head(house_prices)
```




Distributions of prostate cancer variables
===
```{r}
# Get a vector of variable names that are numeric
numeric_vars <- names(house_prices)[sapply(house_prices, is.numeric)]

# Setup grid and margins for plotting
par(mfrow=c(3, 3), mar=c(4, 4, 2, 0.5))

# Loop through each numeric variable and plot its distribution
for (j in 1:length(numeric_vars)) {
  hist(house_prices[, numeric_vars[j]], xlab=numeric_vars[j],
       main=paste("Histogram of", numeric_vars[j]),
       col="lightblue", breaks=20)
}
```

It is apparent that the majority of data has a strong skew and pattern. 

Correlations between House Price variables
===

```{r, include= FALSE}

house_prices_new= select_if(house_prices, is.numeric)
house_prices_new=subset(house_prices_new,select = -c(LotFrontage,MasVnrArea,GarageYrBlt) )
house_prices.cor = cor(house_prices_new)
round(house_prices.cor,3)
```


```{r}

corrplot(cor(house_prices_new[c(1:15)]), type = "full")
corrplot(cor(house_prices_new[c(15:35)]), type = "full")
```

The majority of variables have a moderate relationship between each other. 


Visualizing Relationships Among Variables
===

```{r}
names(house_prices)[sapply(house_prices, is.numeric)]
```


```{r}
#Sale price and General housing information
pairs(~ SalePrice+LotFrontage+LotArea+YearBuilt+YearRemodAdd+GrLivArea, data=house_prices)

#Sale price and Major surface areas
pairs(~ SalePrice+TotalBsmtSF+X1stFlrSF+X2ndFlrSF+TotRmsAbvGrd+GarageArea, data=house_prices)

#Sale price and Basement and "Add-on" areas 
pairs(~ SalePrice+BsmtUnfSF+TotalBsmtSF+GarageArea+WoodDeckSF+OpenPorchSF, data=house_prices)

```

We excluded Variables due to their insignificant information and fixed scales that graphs cannot represent: 

OverallQual, OverallCond, BsmtFinSF1, BsmtFinSF2, MSSubClass, LowQualFinSF, BsmtFullBath, BsmtHalfBath, FullBath, HalfBath, BedroomAbvGr, KitchenAbvGr, Fireplaces, GarageYrBlt, GarageCars, EnclosedPorch, X3SsnPorch, ScreenPorch, PoolArea, MiscVal, MoSold, YrSold


Inspecting relationships over a subset of the observations
===

The majority of variable are evenly distributed and there are only little variables has binary options. For example, 2ndFlrSF can be 0, which indicates the house does not have a second floor. Some house does not have garage, so the area of garage is also 0. Thus, we subset them out and investigate in the rest of data, which contains seocnd floor and a garage. 

```{r}
house_prices_subset = house_prices[house_prices$X2ndFlrSF > 0,]
house_prices_subset = house_prices_subset[house_prices_subset$GarageArea > 0,]
nrow(house_prices_subset)
nrow(house_prices) - nrow(house_prices_subset)

#Sale price and General housing information
pairs(~ SalePrice+LotFrontage+LotArea+YearBuilt+YearRemodAdd+GrLivArea, data=house_prices_subset)

#Sale price and Major surface areas
pairs(~ SalePrice+TotalBsmtSF+X1stFlrSF+X2ndFlrSF+TotRmsAbvGrd+GarageArea, data=house_prices_subset)

#Sale price and Basement and "Add-on" areas 
pairs(~ SalePrice+BsmtUnfSF+TotalBsmtSF+GarageArea+WoodDeckSF+OpenPorchSF, data=house_prices_subset)
```


The general shape is constant comparing with previous one, with half data out, which also proves the evenly distribution of the data.


